Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-11-15 Response-Referees.tex

\textbf{Referee \#6}
\begin{enumerate}
\item \emph{Was the size of the training set of N=250 hourly values of Dst for the GP-AR model (and, additionally, V, Bz, VBz for the GP-ARX model) limited by the numerical effort to perform the matrix inversion? If yes, what resources did the inversion require?}

It is certainly true that Gaussian Process models become prohibitively expensive for very large training sets. Their computational complexity scales as $O(N^3)$, and this was commented in the manuscript on lines $161-165$.

However, a size of 250 is still very much below the computational limits of the method and, in our case, it has been decided by studying the performance of the models for increasing training sets. We have noticed that using sets with more than 250 values does not increase the overall performance.

We have made this point explicit on lines $235-243$: \blockquote{The computational complexity of calculation of the predictive distribution is $O(N^3)$, as discussed in section 2.1. This can limit the size of the covariance matrix constructed from the training data. Note that this computation overhead is paid for every unique assignment to the model hyper-parameters. However, our chosen training set has a size of 250 which is still very much below the computational limits of the method and in our case. Solving equation 13 on a laptop computer takes less than a second for the training set considered in our analysis.

The size of the training set has been decided by studying the performance of the models for increasing training sets. We have noticed that using training sets with more than 250 values does not increase the overall performance in \emph{OSA} prediction of Dst. 
}

@@

Additions were made in lines $229-234$ \blockquote{Although the training and model selection data sets both do not have a geomagnetic storm in them, this would not degrade the performance of \emph{GP-AR} and \emph{GP-ARX} because the linear polynomial kernel describes a non stationary and self similar Gaussian Process. This implies that for two events where the time histories of $Dst$, $V$ and $B_z$ are not close to each other but can be expressed as a diagonal rescaling of time histories observed in the training data, the predictive distribution is a linearly rescaled version of the training data $Dst$ distribution.}

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Model_RMSE_validationStorms}
      \caption{Averaged RMSE performance of \emph{GP-ARX} models for different values of auto-regressive order $p$}
         \label{fig:prmse}
 \end{figure}

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Model_corr_validationStorms}
      \caption{Averaged Correlation Coefficient of \emph{GP-ARX} models for different values of auto-regressive order $p$}
         \label{fig:pcorr}
 \end{figure}

\newpage
\item \emph{The time lag parameters 'p' for the two models were selected by comparing the model to the 63 storm events. Shouldn't the parameter have been selected using an independent set of storms instead in order not to favor these new models against the other models?}

Additions were made in lines $248-262$ in section 4. \blockquote{The time lags chosen for \emph{GP-AR} are $p = 6$ while for \emph{GP-ARX} they are $p=6$, $p_{ex} = 1$. These values are arrived at by building \emph{GP-AR} and \emph{GP-ARX} models for different values of $p$ and choosing the model yielding the best performance on a set of storm events during the periods $1995-1997$ and $2011-2014$.

Box plots \ref{fig:prmse} and \ref{fig:pcorr} show the performance of \emph{GP-ARX} models having values of $p={3,4,5,6,7}$. In these figures lower and upper ends of the vertical lines denote the minimum and maximum respectively, the horizontal line inside a box represents the median, lower and upper ends of a box represent the first and third quartile respectively while dots if present represent outlier points.

We observe that the median \emph{RMSE} and \emph{CC} degrade slightly from $p = 3$ to $p = 7$ and models with $p = 6$ have a much smaller variance in performance making them more robust to outliers. The choice of $p = 6$ thus gives a good trade off between model robustness and complexity.

This approach for choosing the time lags does not lead to bias towards models which only perform well on the data set of \citet{Ji2012} because the storms used to choose the model order have no overlap with it. Moreover, Gaussian Processes are not easily susceptible to overfitting when the number of model hyper-parameters is small.
}

@@

We believe that the referee has misinterpreted the sentence quoted from Ji et al. (2012), and that all of the six models in that paper are indeed run as one-step-ahead models. Our interpretation of that statement is that it refers only to the TL model. That is the only model that indeed does not use the observed DST data as input. It uses some quantities, called $dst*$, that are calculated and advanced in time, by using observed solar wind quantities only. 
However, we agree that that statement is confusing and that it could lead to the referee's interpretation. We present two strong counter-arguments that should settle the issue.

The first argument is that we do not see any reason or benefit for comparing the models in the way the referee suggests they were compared, that is using the predicted $Dst$ values instead of the real and available $Dst$ data. It is obvious that by doing so the performance of the models would degrade, in ways that are difficult to foresee. The main and only purpose of the Ji et al. paper, however, is to compare the performance of different models. Why would one like to compare less-than-optimal results? That paper would be completely flawed if the authors had made what the referee suggests.

We realize that this argument is simply based on scientific common sense, and that the referee might not be persuaded. The following is a more scientific and indisputable proof that the referee's suggestion is wrong.
In the Ji et al. paper, NARMAX and TL are the best performing models. Since the referee's suggestion is irrelevant for TL (because it does not use $Dst$ by construction), we have tested it on NARMAX.
We have taken the NARMAX model as specified in Boynton et al. (2011) and re-tested it on the 63 storms, by using the previously predicted $Dst$ values instead of the actual $Dst$ data. This model is denoted as \emph{NM Variant} in
Figures \ref{fig:rmseNew} and \ref{fig:ccNew}. Here, we show the Root Mean Square Error and the Correlation Coefficient (same as Figure 2b and 2a in Ji et al. (2012)). By using predicted $Dst$, the RMSE changes from 23.4 to 54.8 and the CC goes from 0.87 to 0.04 (which means no-correlation). 
Clearly, the values reported in that paper could not have been obtained in the way the referee suggests.

### 2016-11-15 swsc.tex
Revisions Nov 16

   \abstract{
    We present two models, based on Gaussian Processes, for the \emph{One Step Ahead}(OSA) prediction of the $Dst$ geomagnetic activity index. 
    The models are Auto regressive, with or without eXogenous inputs (GP-ARX and GP-AR, respectively).
    We compare the performance of these models with the current state of the art in one step ahead $Dst$ prediction on a set of 
    63 benchmark storms from 1998-2006, previously analyzed in an earlier study.
    We show that, despite its lack of sophistication, the so-called Persistence model represents an important test to evaluate the performance of a one hour ahead $Dst$ model.
    Contrary to the state-of-the-art models compared in the literature, our models consistently outperform the Persistence model and represent a substantial improvement in the field, when evaluated on standard metrics.
    Finally, an important feature of the new models is that they naturally provide confidence intervals for their forecast.
    }

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

%Vision
In $Dst$ prediction, a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model, which uses the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), the \emph{Persistence} model gives excellent predictive performance on error metrics, despite its apparent lack of sophistication. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction. An obvious critique to the Persistence model is that it is not really predictive, in the sense that it cannot forecast a storm, until it has already commenced. However, most of the literature has used global metrics, such as the Root Mean Square Error (RMSE, see section \ref{sec:exp}) to evaluate models. It is in this context that we argue that the Persistence model should be regarded as the first candidate to outperform.

@@

% Talk about Gaussian Process
We use hourly resolution measurements extracted from NASA/GSFC's OMNI data set through OMNIWeb \citep{OmniPaper}. The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and some technical details about their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res}, bench marked against the \emph{Persistence} as well as the other models outlined in \citet{Ji2012}.

@@

With the multivariate normal distribution defined in equation (\ref{eq:dist}), probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements: $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly, as in equation (\ref{eq:posterior}) (see \citet{Rasmussen:2005:GPM:1162254}).
\begin{equation}
    \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} \sim \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior},
\end{equation}
where
\begin{eqnarray}
    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
\end{eqnarray}

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem $\mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y}$, which can be solved via Cholesky decomposition, with a time complexity of $O(N^3)$, where $N$ is the number of data points.

@@

Below in equations (\ref{eq:Dst}) - (\ref{eq:GPPoly}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$. A vector of features $x_{t-1}$ is used as input to an unknown function $f(x_{t-1})$.

The features $x_{t-1}$ can be any collection of quantities in the hourly resolution OMNI data set. Generally $x_{t-1}$ are time histories of $Dst$ and other important variables such as plasma pressure $p(t)$, solar wind speed $V(t)$, z component of the interplanetary magnetic field $B_z(t)$.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citep{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) z-component ($B_z$) is needed for sub-storms to occur \citep{McPherron1986}. An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citep{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity, IMF $B_z$, and $VB_z$.

@@

The computational complexity of calculation of the predictive distribution is $O(N^3)$ as per section \ref{sec:inference}, this can limit the size of the covariance matrix constructed from the training data. Note that this computation overhead is paid for every unique assignment to the model hyper-parameters. However, our chosen training set has a size of 250 which is still very much below the computational limits of the method and in our case, it has been decided by studying the performance of the models for increasing training sets. We have noticed that using training sets with more than 250 values does not increase the overall performance in \emph{OSA} prediction of Dst.

@@

Box plots \ref{fig:prmse} and \ref{fig:pcorr} show the performance of \emph{GP-ARX} models having values of $p={3,4,5,6,7}$. We observe that the median \emph{RMSE} and \emph{CC} degrade slightly from $p = 3$ to $p = 7$ and models with $p = 6$ have a much smaller variance in performance making them more robust to outliers. The choice of $p = 6$ thus gives a good trade off between model robustness and complexity.

We feel that this approach for choosing the time lags does not lead to bias towards models which only perform well on the data set of \citet{Ji2012} because the storms used have no overlap with it. Also, Gaussian Processes are not easily susceptible to overfitting when the number of model hyper-parameters is small or leave one out cross-validation is not performed.

@@

With respect to the \emph{TL} model one important caveat must be noted. The training of the \emph{TL} model was performed on the OMNI data from 1996-2002 which overlaps with a large number of the storm events, therefore the experimental procedure has a strong bias towards \emph{TL} model.

@@

Figure \ref{fig:cc} shows the comparison of correlation coefficients between model predictions and actual $Dst$ values. From the results of \citet{Ji2012}, the \emph{TL} model gives the highest correlation coefficient of $94\%$, but that is not surprising considering the fact that there is a large overlap between the training data used to train it and storm events. Even so the \emph{GP-ARX} model gives a comparable correlation coefficient to \emph{TL}, although the training set of the \emph{GP-ARX} model has no overlap with the storm events which is not the case for \emph{TL}.

>>>>

   \abstract{
    We present two models, based on Gaussian Processes, for the \emph{One Step Ahead} (OSA) prediction of the $Dst$ geomagnetic activity index. 
    The models are Auto regressive, with or without eXogenous inputs (GP-ARX and GP-AR, respectively).
    We compare the performance of these models with the current state of the art in one step ahead $Dst$ prediction on a set of 
    63 benchmark storms from 1998-2006, previously analyzed in an earlier study.
    We show that, despite its lack of sophistication, the so-called Persistence model represents an important test to evaluate the performance of a one hour ahead $Dst$ model.
    Contrary to the state-of-the-art models compared in the literature, our models consistently outperform the Persistence model and represent a substantial improvement in the field, when evaluated on standard metrics.
    Finally, an important feature of the new models is that they naturally provide confidence intervals for their forecast.
    }

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

%Vision
In $Dst$ prediction, a seemingly trivial yet highly informative prediction method is represented by the so called \emph{Persistence} model, which uses the previous value of $Dst$ as the prediction for the next time step ($\hat{D}st(t) = Dst(t-1)$). Due to high correlation between $Dst$ values one hour apart ($Dst(t)$, $Dst(t-1)$), the \emph{Persistence} model gives excellent predictive performance on error metrics, despite its lack of sophistication. In essence it is a trivial hypothesis and hence it should always be used as a base line to compare the performance of any proposed $Dst$ algorithm. Moreover, given its zero computational cost, a model should at least outperform the \emph{Persistence} model in order to be a viable candidate for advancing the science of space weather prediction. An obvious critique to the Persistence model is that it is not really predictive, in the sense that it cannot forecast a storm, until it has already commenced. However, most of the literature has used global metrics, such as the Root Mean Square Error (RMSE, see section \ref{sec:exp}) to evaluate models. It is in this context that we argue that the Persistence model should be regarded as the first candidate to outperform.

@@

% Talk about Gaussian Process
We use hourly resolution measurements extracted from NASA/GSFC's OMNI data set through OMNIWeb \citep{OmniPaper}. The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and some technical details about their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res}, benchmarked against the \emph{Persistence} as well as the other models outlined in \citet{Ji2012}.

@@

With the multivariate normal distribution defined in equation (\ref{eq:dist}), probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly, as in equation (\ref{eq:posterior}) (see \citet{Rasmussen:2005:GPM:1162254}).
\begin{equation}
    \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} \sim \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior},
\end{equation}
where
\begin{eqnarray}
    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
\end{eqnarray}

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem in Eq. (\ref{eq:posteriormean}), which can be solved via Cholesky decomposition, with a time complexity of $O(N^3)$, where $N$ is the number of data points.

@@

Below in equations (\ref{eq:Dst}) - (\ref{eq:GPPoly}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$. A vector of features $\mathbf{x}_{t-1}$ is used as input to an unknown function $f(\mathbf{x}_{t-1})$.

The features $\mathbf{x}_{t-1}$ can be any collection of quantities in the hourly resolution OMNI data set. Generally $\mathbf{x}_{t-1}$ are time histories of $Dst$ and other important variables such as plasma pressure $p(t)$, solar wind speed $V(t)$, $z$ component of the interplanetary magnetic field $B_z(t)$.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citep{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citep{McPherron1986}. An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citep{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity, IMF $B_z$, and $VB_z$.

@@

The computational complexity of calculation of the predictive distribution is $O(N^3)$, as discussed in section \ref{sec:inference}. This can limit the size of the covariance matrix constructed from the training data. Note that this computation overhead is paid for every unique assignment to the model hyper-parameters. However, our chosen training set has a size of 250 which is still very much below the computational limits of the method and in our case. Solving equation \ref{eq:posteriormean} on a laptop computer takes less than a second for the training set considered in our analysis.

The size of the training set has been decided by studying the performance of the models for increasing training sets. We have noticed that using training sets with more than 250 values does not increase the overall performance in \emph{OSA} prediction of Dst.

@@

Box plots \ref{fig:prmse} and \ref{fig:pcorr} show the performance of \emph{GP-ARX} models having values of $p={3,4,5,6,7}$. In these figures lower and upper ends of the vertical lines denote the minimum and maximum respectively, the horizontal line inside a box represents the median, lower and upper ends of a box represent the first and third quartile respectively while dots if present represent outlier points.

We observe that the median \emph{RMSE} and \emph{CC} degrade slightly from $p = 3$ to $p = 7$ and models with $p = 6$ have a much smaller variance in performance making them more robust to outliers. The choice of $p = 6$ thus gives a good trade off between model robustness and complexity.

This approach for choosing the time lags does not lead to bias towards models which only perform well on the data set of \citet{Ji2012} because the storms used to choose the model order have no overlap with it. Moreover, Gaussian Processes are not easily susceptible to overfitting when the number of model hyper-parameters is small.

@@

With respect to the \emph{TL} model one important caveat must be noted. The training of the \emph{TL} model was performed on data from 1996-2002 which overlaps with a large number of the storm events, therefore the experimental procedure has a strong bias towards \emph{TL} model.

@@

Figure \ref{fig:cc} shows the comparison of correlation coefficients between model predictions and actual $Dst$ values. From the results of \citet{Ji2012}, the \emph{TL} model gives the highest correlation coefficient of $94\%$, but that is not surprising considering the fact that there is a large overlap between the training data used to train it and storm events. Even so the \emph{GP-ARX} model gives a comparable correlation coefficient to \emph{TL}, although the training set of the \emph{GP-ARX} model has no overlap with the storm events, which is not the case for \emph{TL}.
